---
title: "Score-based confidence intervals for random effect variances: FEV1 example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Score-based confidence intervals for random effect variances: FEV1 example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview

This vignette illustrates the `reconf` package using a longitudinal lung function
dataset.  The key feature of the example is that the maximum likelihood estimate
of one variance component falls exactly on the boundary of the parameter space
(zero), a situation where standard Wald and likelihood-ratio confidence intervals
can perform poorly.  Score-based intervals—the default in `reconf`—maintain
correct coverage even at and near such boundary points.

## Data

The `fev1` dataset contains 1994 repeated measurements of forced expiratory
volume in one second (FEV1) for $n = 300$ girls and young women from Topeka,
Kansas, originally described in Fitzmaurice, Laird, and Ware (2011).  Each row
is one observation; the variables are:

| Variable   | Description |
|------------|-------------|
| `id`       | Subject identifier |
| `age`      | Age at measurement (years) |
| `ht`       | Height at measurement (metres) |
| `baseage`  | Baseline age (first visit) |
| `baseht`   | Baseline height (first visit) |
| `logfev1`  | Log FEV1 (natural log, litres) |

```{r load-data}
library(reconf)
library(lme4)

data(fev1)
dim(fev1)
head(fev1)
```

Each subject has between 1 and 12 observations over time.

```{r visits-per-subject}
range(table(fev1$id))
```

## Model

We fit the linear mixed model

$$
\text{logfev1}_{it} = \psi_1 + \psi_2\,\text{age}_{it} + \psi_3\,\text{ht}_{it}
  + \psi_4\,\text{baseage}_i + \psi_5\,\text{baseht}_i
  + U_{i1} + U_{i2}\,\text{age}_{it} + E_{it},
$$

where $U_{i1} \sim N(0,\lambda_1^2)$ and
$U_{i2} \sim N(0,\lambda_2^2)$ are mutually independent random
effects for subject $i$, and $E_{it} \sim N(0,\sigma^2)$ is
measurement error.  The random effects are assumed independent across subjects
and independent of the errors.

The variance parameters of interest are $\lambda = (\lambda_1^2, \lambda_2^2)$;
$\psi$ and $\sigma$ are nuisance parameters.

In `lme4` notation, `(1 | id) + (0 + age | id)` gives independent random
intercepts and random slopes for age, as the model assumes.

```{r fit-model}
fit <- lmer(
  logfev1 ~ age + ht + baseage + baseht + (1 | id) + (0 + age | id),
  data = fev1,
  REML = FALSE
)
```

## Maximum likelihood estimates

```{r mle}
# Fixed effects
round(fixef(fit), 3)

# Variance components
as.data.frame(VarCorr(fit), order = "lower.tri")[
  , c("grp", "var1", "vcov", "sdcor")]
```

The ML estimate $\hat\lambda_1^2 = 0$: the random intercept variance sits exactly
on the boundary of the parameter space.  This is a signal that conventional
asymptotic approximations (Wald and likelihood-ratio) may not be reliable for
$\lambda_1^2$.

## Score-based confidence intervals

`ci_all_lmer()` computes score-based 95% confidence intervals for each variance
component (error variance excluded by default).

```{r ci-all}
ci <- ci_all_lmer(fit)
ci
```

`ci_lmer()` computes the interval for a single parameter and accepts additional
control arguments.  The two parameters here are:

- **Index 1**: $\lambda_1^2$ (random intercept variance, MLE at boundary)
- **Index 2**: $\lambda_2^2$ (random slope variance for age)

```{r ci-single}
ci_lmer(fit, test_idx = 1L)   # random intercept variance
ci_lmer(fit, test_idx = 2L)   # random slope variance
```

## Comparison with likelihood-ratio and Wald intervals

`lme4::confint()` with `method = "profile"` gives profile likelihood-ratio
intervals; Wald intervals are available via `method = "Wald"`.

```{r compare, cache = TRUE}
ci_lr   <- confint(fit, method = "profile", parm = c(".sig01", ".sig02"))
ci_wald <- confint(fit, method = "Wald",    parm = c(".sig01", ".sig02"))
```

```{r compare-table, echo = FALSE}
# confint returns SD-scale bounds; square to get variance scale
tab <- rbind(
  `Score (reconf)`   = ci[1:2, ],
  `Likelihood ratio` = ci_lr^2,
  `Wald`             = ci_wald^2
)
knitr::kable(
  round(tab, 4),
  caption = paste("95% confidence intervals for",
                  "lambda_1^2 (random intercept) and",
                  "lambda_2^2 (random slope)")
)
```

For $\lambda_2^2$ (well away from the boundary) all three methods agree
closely.  For $\lambda_1^2$ (MLE at zero), the score interval is shortest and
its left endpoint is at the boundary, reflecting that the data are consistent
with no random intercept variability.  The likelihood-ratio interval is wider
and the Wald interval is wider still—known deficiencies of those methods near
boundary points of the parameter space.

## Score test for zero random intercept variance

`score_test_lmer()` tests $H_0: \lambda_1^2 = 0$ directly.  We use a REML fit
so that the fixed effects are concentrated out and the null vector only covers
the variance parameters ($\lambda_1^2 = 0$, $\lambda_2^2$, $\sigma^2$).

```{r score-test}
fit_reml <- lmer(
  logfev1 ~ age + ht + baseage + baseht + (1 | id) + (0 + age | id),
  data = fev1, REML = TRUE
)
vc <- as.data.frame(VarCorr(fit_reml), order = "lower.tri")
theta_null_reml <- c(0, vc$vcov[2], sigma(fit_reml)^2)
score_test_lmer(fit_reml, test_idx = 1L, theta_null = theta_null_reml)
```

The large p-value is consistent with the confidence interval: there is little
evidence against $\lambda_1^2 = 0$.

## References

- Fitzmaurice, G. M., Laird, N. M., and Ware, J. H. (2011).
  *Applied Longitudinal Analysis*, 2nd edition. Wiley.
